---
title: "CUNY 607 Project One"
author: "Cassandra Coste"
date: "2/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Load libraries

```{r load libraries and data, message=FALSE, warning=FALSE}

library(tidyverse)
library(knitr)

```



### Read in txt file to clean

```{r}

chess_data <- as.data.frame(read.delim("https://raw.githubusercontent.com/cassandra-coste/CUNY607/main/tournamentinfo.txt", header = FALSE, stringsAsFactors = FALSE, sep = "|"))


```

Remove junk rows (code developed by classmate Gabriel Campos and shared via Slack)

```{r, message=FALSE, warning=FALSE}

toDelete <- seq(1, length(chess_data$V1), 3)
toDelete

chess_data  <- chess_data[-toDelete ,]

```

#### Tidying the dataframe

Pull data for player pre-tournament rating by extracting the digit code following 'R:' 

```{r, message=FALSE, warning=FALSE}
chess_data_clean  <- extract(
  chess_data,
  col=V2,
  into="Player_Rating",
  regex = "(R: +\\d+)", 
  remove = FALSE
)
```

Remove extraneous characters and convert to numeric for player rating column as there is no other pertinent information there

```{r, message=FALSE, warning=FALSE}
chess_data_clean  <- extract(
  chess_data_clean,
  col=Player_Rating,
  into="Player_Rating",
  regex = "(\\d+)", 
  remove = FALSE,
  convert = TRUE
)
```

Extract round data for all 7 rounds removing the original data column 

```{r, message=FALSE, warning=FALSE}
chess_data_clean  <- extract(
  chess_data_clean,
  col = V4,
  into = c("Round_1"),
  regex = "(\\d+)",
  remove = TRUE,
  convert = TRUE
)

chess_data_clean  <- extract(
  chess_data_clean,
  col = V5,
  into = c("Round_2"),
  regex = "(\\d+)",
  remove = TRUE,
  convert = TRUE
)

chess_data_clean  <- extract(
  chess_data_clean,
  col = V6,
  into = c("Round_3"),
  regex = "(\\d+)",
  remove = TRUE,
  convert = TRUE
)

chess_data_clean  <- extract(
  chess_data_clean,
  col = V7,
  into = c("Round_4"),
  regex = "(\\d+)",
  remove = TRUE,
  convert = TRUE
)

chess_data_clean  <- extract(
  chess_data_clean,
  col = V8,
  into = c("Round_5"),
  regex = "(\\d+)",
  remove = TRUE,
  convert = TRUE
)

chess_data_clean  <- extract(
  chess_data_clean,
  col = V9,
  into = c("Round_6"),
  regex = "(\\d+)",
  remove = TRUE,
  convert = TRUE
)

chess_data_clean  <- extract(
  chess_data_clean,
  col = V10,
  into = c("Round_7"),
  regex = "(\\d+)",
  remove = TRUE,
  convert = TRUE
)
```

### Create two data frames containing necessary rows for each column and clean for join

The following steps were employed: 

1) Create indexes of odd and even rows to separate the data into two dataframes containing the pertinent data

2) Remove first rows with messy data

3) Create 'Pair' column in the even dataframe for joining them back together later

4) Renames columns as necessary 

5) Convert the odd dataframe 'Pair' column to numeric for join

6) Select which columns from each dataframe have relevant data


```{r, message=FALSE, warning=FALSE}

even_indexes<-seq(2,130,2)
odd_indexes<-seq(1,129,2)

chess_data_even  <- chess_data_clean[-odd_indexes ,]

chess_data_even = chess_data_even[-1,]

chess_data_even$Pair <- seq.int(nrow(chess_data_even))

chess_data_odd  <- chess_data_clean[-even_indexes ,]

chess_data_odd = chess_data_odd[-1,]

chess_data_odd  <- rename(chess_data_odd, Pair = V1)

chess_data_odd$Pair <- as.integer(chess_data_odd$Pair)

chess_data_odd <- select(chess_data_odd, -c(Player_Rating, V11)) %>% rename(Player_Name = V2) %>% rename(Total_Score = V3)

chess_data_even <- select(chess_data_even, c(V1, Player_Rating, Pair)) %>% rename(Player_State = V1) 

```

### Make final dataframe with full join 

Using a full join, the even and odd dataframes are joined via the 'Pair' column

```{r, message=FALSE, warning=FALSE}

chess_data_final <- full_join(chess_data_even, chess_data_odd, by = "Pair")

```

### Replace pair ID with player ratings for opponent for each round

Create a loop that replaces players Pair ID with their player rating pre-match


```{r, message=FALSE, warning=FALSE}
for (i in 1:nrow(chess_data_final)) {
  for (j in 6:12) {
    if (!is.na(chess_data_final[i,j])) {
      chess_data_final[i,j] <- as.numeric(chess_data_final[chess_data_final[i,j],2])
    }
  }
}
```

### Calculate average opponents rating

```{r, message=FALSE, warning=FALSE}

chess_data_final$Average_Opponent <- rowMeans(chess_data_final[6:12], na.rm = TRUE)
```

### View final table 

```{r, message=FALSE, warning=FALSE}

chess_data_output <- chess_data_final %>%select(Player_Name, Player_State, Total_Score, Player_Rating, Average_Opponent) %>% mutate_if(is.numeric, format, digits=4)

library(knitr) 

kable(chess_data_output, col.names = c('Player Name', 'Player State', 'Total Number of Points', 'Player Pre-Rating', 'Average Rating of Opponents'), align = "ccrrr", caption = "Chess Tournament Scorecard")
```

### Write CSV
 
Code to export final dataframe to CSV

``` {r}

write.csv(chess_data_output,"chess_data_output.csv", row.names = FALSE)
 
