---
title: "CUNY 607 Assignment 3"
author: "Cassandra Coste"
date: "2/15/2021"
output:
   html_document:
    theme: journal

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#### Load libraries

```{r load libraries and data, message=FALSE, warning=FALSE}

library(fivethirtyeight)
library(tidyverse)
library(ggplot2)
```



#### Provide code that identifies the majors that contain either “DATA” or “STATISTICS”

Utilizing the fivethirtyeight College Majors dataset from the following article [https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/],how many times does the word Data or Statistics appear? 

```{r}

college_majors = read.csv('https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv')

sum(str_detect(college_majors$Major, "DATA"))

sum(str_detect(college_majors$Major, "STATISTICS"))

```

Now we know that one major contains the word "DATA" and two majors contain the word "STATISTICS". Next, we can find out which majors these are. 

```{r, message=FALSE, warning=FALSE}

college_majors %>% select("Major") %>% 
  filter(stringr::str_detect(Major, 'DATA|STATISTICS') )

```


#### Write code that transforms the data below:

[1] “bell pepper” “bilberry” “blackberry” “blood orange”

[5] “blueberry” “cantaloupe” “chili pepper” “cloudberry”

[9] “elderberry” “lime” “lychee” “mulberry”

[13] “olive” “salal berry”

Into a format like this:

c(“bell pepper”, “bilberry”, “blackberry”, “blood orange”, “blueberry”, “cantaloupe”, “chili pepper”, “cloudberry”, “elderberry”, “lime”, “lychee”, “mulberry”, “olive”, “salal berry”)


```{r, message=FALSE, warning=FALSE}

fruits_original <- '[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"

[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  

[9] "elderberry"   "lime"         "lychee"       "mulberry"    

[13] "olive"        "salal berry"'

(fruits_original)

fruits_new <- str_extract_all(fruits_original,pattern = '[A-Za-z]+.?[A-Za-z]+')

(fruits_new)

fruits_final <- (fruits_new) %>% 
  str_c(collapse = ", ")

(fruits_final)

writeLines(fruits_final)

```


#### Describe, in words, what these expressions will match:

No. 1 

(.)\1\1

I was not sure if we were suppose to catch something different with the quotation marks missing from No. 1 and 3, but I put this into an R readable format and proceeded, in which case it will match a character repeated three times consecutively in a string. 

```{r, message=FALSE, warning=FALSE}

test1 <- c("Abbb","1234","aaah", "2333")

str_view(test1,"(.)\\1\\1")

```
No. 2  

"(.)(.)\2\1/"

This will match a two consecutive characters followed by them repeating in reverse order within a string.

```{r, message=FALSE, warning=FALSE}

test2 <- c("noon","boop","follow", "2332")

str_view(test2,"(.)(.)\\2\\1")
```


No. 3 

(..)\1

This again was formated for R and in that case it will match a two character set repetition in a string. 


```{r, message=FALSE, warning=FALSE}

test3 <- c("4242","bidi","haha", "onno")

str_view(test3,"(..)\\1")
```


No. 4 

“(.).\1.\1”

This will match a five character set where the first character is repeated as the third and fifth character within the string.

```{r, message=FALSE, warning=FALSE}

test4 <- c("gooog","pepep","12121", "onnno")

str_view(test4,"(.).\\1.\\1")
```
No. 5

"(.)(.)(.).*\3\2\1"

This will match a three consecutive characters repeated in reverse order later within a string, and with any amount of characters between the sets of characters.

```{r, message=FALSE, warning=FALSE}

test5 <- c("abceeeecba","jajaj","haha", "357hello753")

str_view(test5,"(.)(.)(.).*\\3\\2\\1")
```

#### Construct regular expressions to match words that:

No. 1 

Start and end with the same character

```{r, message=FALSE, warning=FALSE}

sample1 <- c("fluff","blob","haha", "tart")

str_view(sample1,"^([A-Za-z]).*\\1$")
```


No. 2 

Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)

```{r, message=FALSE, warning=FALSE}

sample2 <- c("church","blob","haha", "tart")

str_view(sample2,"(..).*\\1")
```

No. 3 

Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)

```{r, message=FALSE, warning=FALSE}

sample3 <- c("eleven","elderberry","haha", "tart")

str_view(sample3,"(.).+\\1.+\\1")
```