---
title: "CUNY 607 Assignment 09"
author: "Cassandra Coste"
date: "4/10/2021"
output:
   html_document:
    theme: journal
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction

The purpose of this assignment is to practice using the NYT's APIs to read data into R and convert to a dataframe. 

### Load libraries

```{r load libraries and data, message=FALSE, warning=FALSE}

library(tidyverse)
library(reactable)
library(jsonlite)

```

### Use NYT API to retrieve movie reviews

For my first attempt to use the API I chose to use the Movie Reviews API to pull the reviews for all the Star Wars movies.


```{r, warning=FALSE, message=FALSE}

# Use NYT API to read in movie reviews for films including star and wars in the names and convert to data frame

nyt_df  <- fromJSON("https://api.nytimes.com/svc/movies/v2/reviews/search.json?query=star+wars&api-key=kMkdHuZBfdpAX7M5j2BUG5oGIP7CmafN", flatten = TRUE) %>% data.frame()

# select columns of interest and clean up column names

nyt_starwars <- nyt_df %>% select(results.display_title, results.mpaa_rating, results.critics_pick, results.byline, results.headline, results.publication_date, results.opening_date) %>% rename(title = results.display_title, rating = results.mpaa_rating, critics_pick = results.critics_pick, byline = results.byline, headline = results.headline, pub_date = results.publication_date, open_date = results.opening_date)


```

### Verify that object is dataframe

Now that I have pulled the data and converted it to a data frame, I perform a check that the conversion from JSON to a data frame was successful. 

```{r, warning=FALSE, message=FALSE}

# check that object we've created is a dataframe 

is.data.frame(nyt_df)

```


### View data frame

Finally, we view the cleaned data frame. 

```{r, warning=FALSE, message=FALSE}
nyt_starwars %>% reactable(bordered = TRUE, striped = TRUE, highlight = TRUE) 
```


### Add additional criteria

Let's say that we only want to view the Star Wars movies that were critic's picks. For this small list, it would be easy enough to eye ball which movies we should watch, but below I practice adding an additional criteria to my API pull to limit my list to only critic picks.

```{r, warning=FALSE, message=FALSE}

# Use NYT API to read in movie reviews for films including star and wars in the names and are also critic's picks convert to data frame

critic_df <- fromJSON("https://api.nytimes.com/svc/movies/v2/reviews/search.json?critics-pick=Y&query=star+wars&api-key=kMkdHuZBfdpAX7M5j2BUG5oGIP7CmafN", flatten = TRUE) %>% data.frame()

# select columns of interest and clean up column names

nyt_starwars_critic <- critic_df %>% select(results.display_title, results.mpaa_rating, results.critics_pick, results.byline, results.headline, results.publication_date, results.opening_date) %>% rename(title = results.display_title, rating = results.mpaa_rating, critics_pick = results.critics_pick, byline = results.byline, headline = results.headline, pub_date = results.publication_date, open_date = results.opening_date)

```

### View data frame

Here we see that if we have limited time, the best of Star Wars films according to the NYT are the three presented below. 

```{r, warning=FALSE, message=FALSE}
nyt_starwars_critic %>% reactable(bordered = TRUE, striped = TRUE, highlight = TRUE) 
```


### Conclusion

The NYT API and supporting documentation is easy to use and can be utilized to pull data for your needs. 


