---
title: "CUNY 607 Assignment 05"
author: "Cassandra Coste"
date: "3/6/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries

```{r load libraries and data, message=FALSE, warning=FALSE}

library(tidyverse)
library(knitr)

```

### Read in csv file to clean

```{r}

flights_raw <- as.data.frame(read.delim("https://github.com/cassandra-coste/CUNY607/raw/main/Flights%20Data.csv", header = TRUE, stringsAsFactors = FALSE, na.strings=c("","NA"), sep = ",", fileEncoding = "UTF-8-BOM"))

```

#### Tidying the dataframe

```{r, message=FALSE, warning=FALSE}

# rename nonsensical columns and use fill to fill missing airline data from Airline column

flights_clean <- flights_raw %>% rename(Airline = X, Status = X.1) %>% fill(Airline)

# delete row 3 which contains no data

flights_clean <- flights_clean [-3,]

# gather data using City as the key to separate city variable from count data on on-time versus delayed flights 

flights_clean <- gather(flights_clean, City, Count, 3:7)

# covert Count column to numeric accounting for untidy numbers that came with CSV file 

flights_clean[, 4]  <- as.numeric(gsub(",","",flights_clean[, 4]))

# use spread to separate out on-time variable from delayed variable

flights_clean <- flights_clean %>% spread(Status, Count)

# rename delayed and on-time columns and replace period with space in city column

flights_clean <- flights_clean %>% rename(Delayed = delayed, On_time = "on time")

flights_clean$City <- str_replace(flights_clean$City,"\\."," ")
```

### View tidy data

```{r, message=FALSE, warning=FALSE}

kable(flights_clean , col.names = c('Airline', 'Departing City', 'Delayed', 'On Time'), align = "ccrr", caption = "Flights Data by City")
```

### Visualize Data

```{r, message=FALSE, warning=FALSE}

# Compute percentage of flights delayed

flights_Visualize <- flights_clean %>% mutate(total = rowSums(across(where(is.numeric)))) %>% mutate(percent_delayed = Delayed/total * 100)

## Look at percentage of flights delayed by airline by city

ggplot(flights_Visualize, aes(fill = Airline, x = reorder(City,-percent_delayed), y = percent_delayed)) + geom_bar(position = "dodge", stat="identity") + coord_flip() + ggtitle("Percentage of Flights Delayed by Airline") + ylab("Percentage of Flights Delayed") + xlab("Departing City") + scale_fill_discrete(name = "Airline", labels = c("Alaska", "America West"))
```

#### Takeaway 

After interpreting this visualization, it's easier to see that across the board, America West airlines experiences a higher percentage of flights delayed than Alaska airlines across all cities evaluated. You can also see that certain cities experience a greater number of delays all around for both airlines than other cities (i.e. if you are flying out of Phoenix on Alaska or America West Airlines, you might not need to worry about delays are much as if you are flying out of San Francisco). 
